---
title: "Data Mining - Clustering"
output: html_notebook
---
## Data loading
```{r}
install.packages('ggplot2')
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
preprocess_data = function(path){
  original = as.data.frame(read.table(path, header=F))
  labels = as.array(original[,ncol(original)])+1
  original[,ncol(original)] = NULL
  scaled = as.data.frame(scale(original))
  dataset = list(original=original, labels=labels, scaled=scaled)
  return(dataset)
}

```

## Algorithms

### Hierarchical Clustering
```{r}
hierarchical = function(dataset, method, title) {
  dist_mat = dist(dataset$scaled, method = 'euclidean')
  hclust_avg = hclust(dist_mat, method = method)
  cut_avg = cutree(hclust_avg, k = length(levels(as.factor(dataset$labels))))
  dataset$cluster = cut_avg
  ari = ARI(dataset$labels, dataset$cluster)
  ggplot(dataset$original, aes(x=V1, y=V2, color = factor(dataset$cluster))) + geom_point() + ggtitle(title, paste("ARI:", ari)) + labs(color = "Clusters")
}

```

### k-Means
```{r}
k_means = function(dataset) {
  set.seed(20)
  kmeans_clusters = kmeans(dataset$scaled, length(levels(as.factor(dataset$labels))))
  kcl = as.factor(kmeans_clusters$cluster)
  ari = ARI(dataset$labels, kcl)
  ggplot(dataset$original, aes(x=V1, y=V2, color = factor(kcl))) + geom_point() + ggtitle("k-Means", paste("ARI:", ari))  + labs(color = "Clusters")
}

```

### DBSCAN
```{r}
install.packages("dbscan")
suppressPackageStartupMessages(library(dbscan))

db_scan = function(dataset, eps, k) {
  dbscan_cluster = dbscan(dataset$original, eps=eps, minPts=k)
  dcl = dbscan_cluster$cluster
  ari = ARI(dataset$labels, dcl)
  #cat(length(levels(as.factor(dbscan_cluster$cluster))), "ARI:",ari,eps,k,"\n")
  ggplot(dataset$original, aes(x=V1, y=V2, color = factor(dcl))) + geom_point() + ggtitle("DBSCAN", paste("ARI:", ari))  + labs(color = "Clusters")
}

```

### ARI
```{r}
install.packages("MixGHD", dependencies=TRUE, repos='http://cran.rstudio.com/')
suppressPackageStartupMessages(library(MixGHD))
```


### 2d10c dataset
Best algorithm: DBSCAN

The clusters have density differences but given their spatial remoteness, with some hyperparameter fine-tuning, most of the elements of the sparse clusters should be able to be clumped together while still preventing the merge of the close, dense clusters
```{r}
data = preprocess_data('datasets/2d-10c.dat')
ari = ARI(data$labels, data$labels)
ggplot(data$original, aes(x=V1, y=V2, color = factor(data$labels))) + geom_point() + ggtitle("Original", paste("ARI:", ari))  + labs(color = "Clusters")
hierarchical(data, 'single', 'Single linkage')
hierarchical(data, 'complete', 'Complete linkage')
hierarchical(data, 'average', 'Average linkage')
hierarchical(data, 'ward.D', 'Ward variance')
k_means(data)
db_scan(data, eps=0.8, k=13)
#Parameters chosen via grid search

```


### Smile dataset

Best algorithm: single-linkage/DBSCAN

The clusters are well spread apart and relatively dense, with elongated shapes, therefore the two algorithms should have no issues with them
```{r}
data = preprocess_data('datasets/smile.dat')
ari = ARI(data$labels, data$labels)
ggplot(data$original, aes(x=V1, y=V2, color = factor(data$labels))) + geom_point() + ggtitle("Original", paste("ARI:", ari))  + labs(color = "Clusters")
hierarchical(data, 'single', 'Single linkage')
hierarchical(data, 'complete', 'Complete linkage')
hierarchical(data, 'average', 'Average linkage')
hierarchical(data, 'ward.D', 'Ward variance')
k_means(data)
db_scan(data, eps = 0.6, k = 5)

```


### Iris dataset
Best algorithm: DBSCAN(?)

The data is 4d which makes it difficult to visualize and thus predict the best algorithm. PCA shows that the first component accounts for >90% of the variation in the data, so the clusters should be in some way both elongated and rounded, so any algorithm has a chance.
```{r}
data = preprocess_data('datasets/iris.dat')
ari = ARI(data$labels, data$labels)
ggplot(data$original, aes(x=V1, y=V2, color = factor(data$labels))) + geom_point() + ggtitle("Original", paste("ARI:", ari))  + labs(color = "Clusters")
hierarchical(data, 'single', 'Single linkage')
hierarchical(data, 'complete', 'Complete linkage')
hierarchical(data, 'average', 'Average linkage')
hierarchical(data, 'ward.D', 'Ward variance')
k_means(data)
db_scan(data, eps = 0.4, k = 4)
# Parameters chosen via grid search
```

### Long dataset

Best fit: single-linkage (elongated clusters)
```{r}
data = preprocess_data('datasets/long.dat')
ari = ARI(data$labels, data$labels)
ggplot(data$original, aes(x=V1, y=V2, color = factor(data$labels))) + geom_point() + ggtitle("Original", paste("ARI:", ari))  + labs(color = "Clusters")
hierarchical(data, 'single', 'Single linkage')
hierarchical(data, 'complete', 'Complete linkage')
hierarchical(data, 'average', 'Average linkage')
hierarchical(data, 'ward.D', 'Ward variance')
k_means(data)
db_scan(data, eps = 0.4, k = 2)

```

### Order2-3clust dataset

Best algorithm: single-linkage (elongated clusters)

DBSCAN might perform well but the difference between the clusters' densities seems too high
```{r}
data = preprocess_data('datasets/order2-3clust.dat')
ari = ARI(data$labels, data$labels)
ggplot(data$original, aes(x=V1, y=V2, color = factor(data$labels))) + geom_point() + ggtitle("Original", paste("ARI:", ari))  + labs(color = "Clusters")
hierarchical(data, 'single', 'Single linkage')
hierarchical(data, 'complete', 'Complete linkage')
hierarchical(data, 'average', 'Average linkage')
hierarchical(data, 'ward.D', 'Ward variance')
k_means(data)
db_scan(data, eps = 2, k = 7)
```


### Square dataset

Best algorithm: k-Means/complete-linkage (round clusters)

Likely no algorithm performs well, as the data is generated from a randomly split uniform distribution with no clear boundaries
```{r}
data = preprocess_data('datasets/square.dat')
ari = ARI(data$labels, data$labels)
ggplot(data$original, aes(x=V1, y=V2, color = factor(data$labels))) + geom_point() + ggtitle("Original", paste("ARI:", ari))  + labs(color = "Clusters")
hierarchical(data, 'single', 'Single linkage')
hierarchical(data, 'complete', 'Complete linkage')
hierarchical(data, 'average', 'Average linkage')
hierarchical(data, 'ward.D', 'Ward variance')
k_means(data)
db_scan(data, eps = 0.7, k = 10)

```



































